---
title: "HMDA Analysis"
output: pdf_document
date: "2024-02-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages
```{r}
library(ggplot2)
library(dplyr)
library(rvest)
library(plm)
library(sjPlot)
library(sjmisc)
library(sjlabelled)
```
## I) Literature Review

# This article gives a good intro to what makes a person more likely to be approved for a mortgage: https://themortgagereports.com/47861/what-are-your-chances-of-mortgage-approval-down-payment-credit-score

# Summary of this article - the 3 factors which make a person more likely to be approved for a mortgage are - a) income and debts, b) credit score, and c) assets. When a person is strong in all 3 of these categories, then they are more likely to be approved for a mortgage. In our data set, we don't have information on credit score and assets, but we know the applicant's income. Stable job is anonther important factor (we have data for unemployment rate). It also says that FHA and VA (Veteran) loans are much more leniant on credit scores (this is under the variable, loan_type_name, in our data set) so this signifies the relevance of having loan_type_name as a control in our model. 

# To our original data set (https://www.consumerfinance.gov/data-research/hmda/), we have added the following new variables - 
# a) unemployment rate - https://www.bls.gov/web/laus/laumstrk.htm
# b) CPI-U - https://www.usinflationcalculator.com/inflation/consumer-price-index-and-annual-percent-changes-from-1913-to-2008/#google_vignette
# c) Per Capita Real GDP - https://apps.bea.gov/regional/histdata/releases/0615gsp/index.cfm

# We got the HMDA data set which basically is a regulation requiring financial institutions to report their mortage applicant's characteristics such as demographic characteristics (race, gender, state of application) and economic indicators (applicant income, loan amount obtained and so on) to ensure that financial instituions don't indulge in discrimantory practices when issuing mortgages. This is in allignment with the Fair Housing Act (1968) (read this article - https://www.hud.gov/program_offices/fair_housing_equal_opp/fair_lending#:~:text=The%20Fair%20Housing%20Act%20makes,Approvals%20and%20denials) which makes it illegal to discriminate against someone because of race, color, religion, sex (including gender, gender identity, sexual orientation, and sexual harassment), familial status, national origin or disability at any stage of the mortgage process, including Approvals and denials, terms like interest rates, points, fees and other costs, advertising, mortgage broker services, property appraisal, servicing, home loan modification assistance, and homeowners insurance. In our questions, we are basically focusing on the race and gender aspect of the applicants to see if any discriminattion still happens despite these regulations. Our pursuit of this question is shaped by the following literature review - 

# 1) Bias in the mortgage approval process (https://www.investopedia.com/bias-in-the-mortgage-approval-process-5181654) - Historically, homeownership has been influenced race, ethnicity, and other prejudices. Here, you can give an example of red lining used to be a discrimatory practice in homeowenership in the US. This has been a major problem since for many families, their home is the single biggest asset and represents a significant portion of their total wealth - this particularly applies for non-white homeowners. Given the importance ofb housing as means of survival along with being a major asset, it is also important to recogonize that most homeowners can't afford to purchase a home outright, implying the significance of the mortgage/lending industry, which is seen substantial growth over the past. Nonetheless, despite regulations such as the Fair Hosuing ACt of 1968, there continues to be discrimination in this sector of mortgage approvals. In fact, a comprehensiver review of the evidence published by the urabn institute in 1999 found that minority homeowners faced discrimination. You can read more about the report here and summarize its results if possible (chrome-extension://efaidnbmnnnibpcajpcglclefindmkaj/https://www.urban.org/sites/default/files/publication/60776/410821-Discrimination-in-Metropolitan-Housing-Markets.PDF). In fact, over 74.5% of homeowners in 2nd quarter of 2022 were non-hispnaic white americans as compared to 45.3% Black americans and 57.3% of all other races such as (Asians, Native Hawaiians, and so on). Note that asians also are associated with lesser homeownership - we can also explore the racial gap in mortgage approval for asians and white applicants. Given these facts, we are interested in focusing on the race aspect of discriminatory practices.

# 2) Gender Discrimination in US Mortgage Lending: A Temporal and Spatial Analysis (chrome-extension://efaidnbmnnnibpcajpcglclefindmkaj/https://repository.library.georgetown.edu/bitstream/handle/10822/1052744/Honors%20%28Final%29%20-%20Fatma%20Marafi.pdf?sequence=1&isAllowed=y) - This research paper talks about our reasoning behind focusing not only race but also the gender aspect of lending practices. It shows that women face a higher denial rate even after several regulations - you can raed more about this. Along with this two very important findings of this paper are discrimination is 1) pro-cyclical (i.e. peaks in recessions) and that  2) it varies substantially across states.  Specifically, gender discrimination in lending is higher in states with (i) high level of conservatism, (ii) low support for gay rights, and (iii) low female representation in the state legislature. These 2 findings (1 and 2) provide support for why we are interested in the period aspect of this study (recession vs recovery) and moreover, it leads us into our other question where we see if black applicants have a higher approval rate in California than Virginia in recession and recovery periods since California is generally considered to be the more liberal state than VA. 

# 3) Racial Segregation and the American Foreclosure Crisis (https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4193596/#:~:text=Compared%20to%20whites%20with%20similar,et%20al%202005%2C%202007).) - this research paper gives a very important finding during that during housing bubble, before the great recessioin (2007 - 2009) which saw a rise in subprime mortgages, African American and Latino homeowners bore a disproportionate share of costs stemming from the housing bubble. Compared to whites with similar credit profiles, down payment ratios, personal characteristics, and residential locations, African Americans were much more likely to receive subprime loans. This means that even though this increases the likelihood for black applicants to be approved for a mortgage in the bubble period, they become vulnerable to the several disadvantages associated subprime loans which are the ffollowing - 

#i)Subprime mortgages are considered bad because they are loans offered to individuals with poor credit histories or other risk factors that make them more likely to default on their mortgage payments. These loans typically come with higher interest rates and fees compared to prime mortgages, which are offered to borrowers with strong credit profiles.

#ii)The disproportionate impact of subprime lending on African American and Latino homeowners during the housing bubble can be attributed to several factors:

#iii)Predatory Lending Practices: Subprime lenders often engage in predatory lending practices, such as targeting minority communities with deceptive marketing tactics, steering borrowers into loans they cannot afford, and charging excessive fees and interest rates.

#iv)Systemic Discrimination: Discriminatory lending practices have historically disadvantaged minority communities. Even when African American and Latino borrowers have similar credit profiles and financial characteristics as their white counterparts, they may still be steered towards subprime loans due to systemic biases and institutional racism within the lending industry.

#v)Limited Access to Traditional Credit: Many African American and Latino households have limited access to traditional banking services and may be more reliant on subprime lenders for financing. This limited access to mainstream financial institutions can leave them vulnerable to predatory lending practices.

#vi)Community Disinvestment: Historically, minority communities have faced disinvestment and neglect from financial institutions and policymakers, resulting in lower property values and fewer opportunities for wealth accumulation. Subprime lending can exacerbate these disparities by further exploiting the financial vulnerability of these communities.

#vii)Overall, the disproportionate impact of subprime lending on African American and Latino homeowners reflects broader systemic inequalities within the housing and financial sectors. Subprime mortgages are considered bad because they often exploit vulnerable borrowers, contribute to financial instability, and perpetuate cycles of poverty and inequality.

# To conclude the suprime mortgages make the black applicants more likely to be approved for a mortgage but increase their vulnerability to default on the loan. (In fact our result is in allignment with this since we find that the racial gap between white and balck female applicants is lesser during recession than recovery but this may be a direct consequence of easy accessibilty to subprime mortgages which isn't neccessarily good).


# Data Prep - data cleaning of the original random sample of 50,000 observations and adding new variables for CPI-U, Unemployment Rate, and 
# Per Capita Real GDP
```{r}
# Load in HMDA Data
HMDA<-read.csv("final_data_HMDA.csv")

# Converting to factor variables
HMDA$state_abbr <- factor(HMDA$state_abbr)
HMDA$agency_abbr <- factor(HMDA$agency_abbr)
HMDA$loan_type_name <- factor(HMDA$loan_type_name)
HMDA$property_type_name <- factor(HMDA$property_type_name)
HMDA$loan_purpose_name <- factor(HMDA$loan_purpose_name)
HMDA$preapproval_name <- factor(HMDA$preapproval_name)
HMDA$action_taken_name <- factor(HMDA$action_taken_name)
HMDA$applicant_race_name_1 <- factor(HMDA$applicant_race_name_1)
HMDA$applicant_sex_name <- factor(HMDA$applicant_sex_name)

# Reducing applicant_race_name_1 to only 2 levels - "White", "Black", "Asian"
HMDA <- mutate(HMDA,
              app_race = case_when(
                applicant_race_name_1 %in% c("Asian") ~ "Asian",
                applicant_race_name_1 %in% c("Black or African American") ~ "Black",
                applicant_race_name_1 %in% c("White") ~ "White",
                TRUE ~ NA_character_
              )
)

# Converting app_race to a factor variable
HMDA$app_race <- factor(HMDA$app_race)


# Reducing applicant_sex_name to Male and Female
HMDA$applicant_sex_name <- factor(
  HMDA$applicant_sex_name,
  levels = c("Male", "Female")
)

# Converting applicant_sex_name to a factor variable
HMDA$applicant_sex_name <- factor(HMDA$applicant_sex_name)


# Create unemployment rates column
unemployment_rates <- data.frame(
  state_abbr = c(rep('VA', 6), rep('CA', 6)),
  Year = rep(2007:2012, 2),
  unemployment_rate = c(3.0, 3.9, 6.7, 6.9, 6.4, 5.9, 5.4, 7.2, 11.3, 12.4, 11.8, 10.4)
)

# Join the unemployment rates with the HMDA data
HMDA <- merge(HMDA, unemployment_rates, by = c("state_abbr", "Year"), all.x = TRUE)

# Adding a column for per capita real GDP for VA and CA (state-level)
HMDA <- HMDA %>%
  mutate(
    Per_Cap_GDP = case_when(
      state_abbr == "VA" & Year == 2007 ~ 52676,
      state_abbr == "VA" & Year == 2008 ~ 52140,
      state_abbr == "VA" & Year == 2009 ~ 51772,
      state_abbr == "VA" & Year == 2010 ~ 52370,
      state_abbr == "VA" & Year == 2011 ~ 52145,
      state_abbr == "VA" & Year == 2012 ~ 51967,
      state_abbr == "CA" & Year == 2007 ~ 55011,
      state_abbr == "CA" & Year == 2008 ~ 54733,
      state_abbr == "CA" & Year == 2009 ~ 51830,
      state_abbr == "CA" & Year == 2010 ~ 51878,
      state_abbr == "CA" & Year == 2011 ~ 52037,
      state_abbr == "CA" & Year == 2012 ~ 52851,
      TRUE ~ NA_integer_  # for any other cases
    )
  )

# Adding a column for average annual CPI-U (National level)
HMDA <- HMDA %>%
  mutate(
    CPI = case_when(
      Year == 2007 ~ 207.3,
      Year == 2008 ~ 215.3,
      Year == 2009 ~ 214.5,
      Year == 2010 ~ 218,
      Year == 2011 ~ 225,
      Year == 2012 ~ 230,
      TRUE ~ NA_integer_ 
    )
  )

# Removing unwanted columns
HMDA <- select(HMDA, -agency_abbr, -owner_occupancy_name, -preapproval_name, 
               -co_applicant_race_name_1, -co_applicant_sex_name, -purchaser_type_name, 
               -hud_median_family_income, -hoepa_status_name, -census_tract_number, -applicant_ethnicity_name,
               -co_applicant_ethnicity_name)

# Adding a new column for loan_amount_000s / applicant_income_000s

HMDA$loan_to_income <- HMDA$loan_amount_000s / HMDA$applicant_income_000s

# Renaming the level of "One-to-four family dwelling" to "Traditional housing" in the variable, property_type_name
levels(HMDA$property_type_name)[levels(HMDA$property_type_name) == "One-to-four family dwelling (other than manufactured housing)"] <- "Traditional housing"

# Removing na values
HMDA <- na.omit(HMDA)

# Rempoving outliers for loan_to_income
# Calculate the IQR for loan_to_income
IQR_loan_to_income <- IQR(HMDA$loan_to_income)

# Calculate the first and third quartiles
Q1 <- quantile(HMDA$loan_to_income, 0.25)
Q3 <- quantile(HMDA$loan_to_income, 0.75)

# Define the lower and upper bounds for what's considered non-outlier data
lower_bound <- Q1 - 1.5 * IQR_loan_to_income
upper_bound <- Q3 + 1.5 * IQR_loan_to_income

# Filter the dataset to exclude outliers
HMDA <- HMDA %>%
  filter(loan_to_income >= lower_bound & loan_to_income <= upper_bound)

summary(HMDA)
```
## EDA ##

#Graph to Compare Genders and Race Simultaneously using a bar plot (for both recession and recovery)
```{r}
# Calculate approval rate
approval_data <- HMDA %>%
  group_by(state_abbr, app_race, applicant_sex_name, action_taken_name) %>%
  summarize(count = n()) %>%
  tidyr::spread(action_taken_name, count, fill = 0) %>%
  mutate(approval_rate = `Loan originated` / (`Loan originated` + `Loan denied`) * 100)

# First, calculate the mean approval rate for each group
approval_data_summary <- approval_data %>%
  group_by(app_race, applicant_sex_name) %>%
  summarize(mean_approval_rate = mean(approval_rate, na.rm = TRUE)) %>%
  ungroup()

# Now create the stacked bar plot
ggplot(approval_data_summary, aes(x = app_race, y = mean_approval_rate, fill = applicant_sex_name)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Mean Approval Rate by Race and Applicant Sex",
       x = "Race",
       y = "Mean Approval Rate (%)", fill = "Sex") +
  scale_fill_manual(values = c("Male" = "blue", "Female" = "red"))
```
Women have lower approval rate for all 3 races - difference seems highest for black applicants, then white, and least for asian

Therefore, we focus only on Female applicants in the following sections.

Now, creating a bar plot to compare mean approval rate of female and male applicants across the 2 periods (Recession and Recovery) - 

```{r}
# Calculate approval rate
approval_data <- HMDA %>%
  group_by(state_abbr, app_race, action_taken_name, period, applicant_sex_name) %>%
  summarize(count = n()) %>%
  tidyr::spread(action_taken_name, count, fill = 0) %>%
  mutate(approval_rate = `Loan originated` / (`Loan originated` + `Loan denied`) * 100)

# First, calculate the mean approval rate for each group
approval_data_summary <- approval_data %>%
  group_by(period, applicant_sex_name) %>%
  summarize(mean_approval_rate = mean(approval_rate, na.rm = TRUE)) %>%
  ungroup()

# Now create the stacked bar plot
ggplot(approval_data_summary, aes(x = period, y = mean_approval_rate, fill = applicant_sex_name)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Mean Approval Rate by Period and Applicant Sex",
       x = "Period (Recession/Recovery)",
       y = "Mean Approval Rate (%)", fill = "Sex") +
  scale_fill_manual(values = c("Male" = "blue", "Female" = "red"))
```
Furthermore, we can view that the gender gap increases during the cyclical fluctuations such as the period of The Great Recession. Here, even though the approval rate increased for both male and female applicants during the recovery period, the difference in approval rate between male and female applicants increased, implying that females were even worse off during the recovery period than before. 

Therefore, we then perform EDA to view which factors influence female applicant's loan approval status during the recovery period and the recession period - this way, we view how these influences on female approval rate differ across the 2 periods of Recovery and Recession

First, we start with the recovery period

## Subsetting our data for females and recovery
```{r}
# Step 1: Filter the data for female applicants and the Recovery period
HMDA_female_Recovery <- HMDA %>%
 filter(period == "Recovery" & applicant_sex_name == "Female")
```

## EDA for the subsetted data (female applicants and recovery period) - Note: also refer to EDA for our random sample of 50,000 observations we did before (it's on the drive under the EDA folders)
```{r}
#EDA - for female applicants and the Recovery period (Interactions)
#1) summary table
summary(HMDA_female_Recovery)
# action_taken_name is a balanced data set 

# exploring the relation for interactions that I have added

#i) interaction between loan_to_income and property_type_name
ggplot(HMDA_female_Recovery, aes(x = property_type_name, y = loan_to_income, fill = action_taken_name)) +
  geom_violin(scale = "width", trim = FALSE) +
  labs(title = "Female Loan to Income Ratio by Property Type and Loan Status (Recovery)",
       x = "Property Type",
       y = "Loan to Income Ratio",
       fill = "Loan Status") +
  theme_minimal() +
  scale_fill_manual(values = c("Loan originated" = "blue", "Loan denied" = "red"))


##ii) interaction between loan_to_income and state_abbr
ggplot(HMDA_female_Recovery, aes(x = state_abbr, y = loan_to_income, fill = action_taken_name)) +
  geom_violin(scale = "width", trim = FALSE) +
  labs(title = "Female Loan to Income Ratio by State and Loan Status (Recovery)",
       x = "State",
       y = "Loan to Income Ratio",
       fill = "Loan Status") +
  theme_minimal() +
  scale_fill_manual(values = c("Loan originated" = "blue", "Loan denied" = "red"))

##iii) interaction between loan_to_income and loan_purpose_name
ggplot(HMDA_female_Recovery, aes(x = loan_purpose_name, y = loan_to_income, fill = action_taken_name)) +
  geom_violin(scale = "width", trim = FALSE) +
  labs(title = "Female Loan to Income Ratio by Loan Purpose and Loan Status (Recovery)",
       x = "Loan Purpose",
       y = "Loan to Income Ratio",
       fill = "Loan Status") +
  theme_minimal() +
  scale_fill_manual(values = c("Loan originated" = "blue", "Loan denied" = "red"))

##iv) interaction between per capita GDP and CPI
ggplot(HMDA_female_Recovery, aes(x = Per_Cap_GDP, y = CPI, color = action_taken_name)) +
  geom_smooth(method = "lm", se = F) +
  labs(title = "State-Level Per Capita GDP vs. CPI (Recovery)",
       x = "Per Capita GDP",
       y = "CPI",
       color = "Loan Status") +
  scale_fill_manual(values = c("Loan originated" = "blue", "Loan denied" = "red")) +
  theme_minimal()
```
Interpretation of EDA for Interactions - 

i) Interaction between Loan to income ratio and property type - Traditional housing mortgage application exhibit a higher loan to income ratio as compared to manufactured housing. This result is in aligned with expectations since traditional housing options are much more expensive and entail a higher loan amount as compared to manufactured housing options, i.e. the amount of loan applied/obtained for traditional housing will be higher than manufactured housing for the same level of applicant's income. Moreover, in the case of loan denials, the distribution of loan to income ratio for traditional housing seems to be slightly more left skewed and broader at higher values of loan to income ratio as compared to the distribution for manufactured housing. This means that more denials are experienced for higher values of loan to income ratio in the case of traditional housing as compared to manufactured housing, indicating a possible interaction between property type and loan to income ratio.

ii) Interaction between Loan to Income Ratio and State - We can observe that the distribution of loan to income ratio is similar for loan denials and loan originations in the case of California. This finding is in alignment with the fact that California adopted a liberal loan approval policy with few barriers to getting a loan approved, and a higher loan to income ratio didn't have a significant impact on whether a loan was approved or denied in California. However, the left skewed distribution of loan to income ratio, in the case of denials, for Virginia shows that applicants with a lower creditworthiness were more likely to be denied a loan. This differs from Virginia's distribution of loan to income ratio in the case of loan approvals since the distribution is right skewed and shows that applicants with a lower loan to income ratio, and therefore a higher creditworthiness, were more likely to be approve for a loan. These results indicate a possible interaction between loan to income ratio and state.


iii) Interaction between Loan to income ratio and Loan Purpose - We can observe that loans for home improvements are associated with a lower loan to income ratio as compared to those meant for home purchase or refinancing since home improvement loans are usually smaller loan amounts meant for minor changes. Moreover, in the case of denials, the distribution of loan to income ratio for home purchases is slightly more left skewed as compared to loans for refinancing which indicates that a higher loan to income ratio corresponds to more denials in the case of home purchases as compared to refinancing. This indicates a possible interaction between loan to income ratio and loan purpose.


iv) Interaction between per capita GDP and CPI - The plot shows that a higher state-level per capita GDP is associated with a higher state-level CPI, which is alignment with the fact that as people's standards of living increase (as indicated by the per capita GDP), it results in a higher CPI due to inflation and a greater demand for goods and services. Moreover, it can be observed an intersection of slopes, with the slope for loan denials being steeper than loan originations, which implies that a higher per capita GDP in a state may not necessarily result in more loan originations if there's inflation in housing prices which neutralize the effect of a higher applicant income. This suggests a possible interaction between per capita GDP and CPI. 

v) State and Approval Rate

```{r}
# Calculate approval rate
approval_data_recovery <- HMDA_female_Recovery %>%
  group_by(state_abbr, loan_type_name, app_race, action_taken_name) %>%
  summarize(count = n()) %>%
  tidyr::spread(action_taken_name, count, fill = 0) %>%
  mutate(approval_rate = `Loan originated` / (`Loan originated` + `Loan denied`) * 100)

approval_data_summary <- approval_data_recovery %>%
  group_by(state_abbr) %>%
  summarize(mean_approval_rate = mean(approval_rate, na.rm = TRUE))

ggplot(approval_data_summary, aes(x=state_abbr, y=mean_approval_rate)) +
  geom_bar(stat="identity", fill="blue") +
  labs(title="Female Mean Approval Rate by State (Recovery)",
       x="State", 
       y="Mean Approval Rate (%)") + 
  theme_minimal() +
  scale_y_continuous(limits = c(0,60,20))
```
VA has a higher approval rate than CA during recovery!

vi) Loan Type and Approval Rate
```{r}
approval_data_summary <- approval_data_recovery %>%
  group_by(loan_type_name) %>%
  summarize(mean_approval_rate = mean(approval_rate, na.rm = TRUE))

ggplot(approval_data_summary, aes(x=loan_type_name, y=mean_approval_rate)) +
  geom_bar(stat="identity", fill="blue") +
  labs(title="Female Mean Approval Rate by Loan Type (Recovery)",
       x="Loan Type", 
       y="Mean Approval Rate (%)") + 
  theme_minimal() +
  scale_y_continuous(limits = c(0,60,20))
```
FSA/RHS-guaranteed > Conventional > FHA-insured > VA-guaranteed

vii) Race and Approval Rate
```{r}
approval_data_summary <- approval_data_recovery %>%
  group_by(app_race) %>%
  summarize(mean_approval_rate = mean(approval_rate, na.rm = TRUE))

ggplot(approval_data_summary, aes(x=app_race, y=mean_approval_rate)) +
  geom_bar(stat="identity", fill="blue") +
  labs(title="Female Mean Approval Rate by Race (Recovery)",
       x="Race", 
       y="Mean Approval Rate (%)") + 
  theme_minimal() +
  scale_y_continuous(limits = c(0,80,20))
```
White > Asian > Black

## i) Logistic Regression for the Recovery Period

```{r}
# Step 3: Logistic Regression

# Having Black as the reference class
HMDA_female_Recovery$app_race <- relevel(HMDA_female_Recovery$app_race, ref = "Black")

# Fit a logistic regression model with "action_taken_name" as the response variable
model <- glm(action_taken_name ~ state_abbr + loan_type_name + property_type_name +
             loan_purpose_name + loan_to_income + app_race + CPI + unemployment_rate + Per_Cap_GDP + loan_to_income*property_type_name + loan_to_income*state_abbr + loan_to_income * loan_purpose_name + Per_Cap_GDP * CPI, data = HMDA_female_Recovery, family = binomial)

# Step 4: Hypothesis Testing
# Get a summary of the model to check p-values
summary(model)

##set seed so results are reproducible
set.seed(111)

##evenly split data into train and test sets
sample.data<-sample.int(nrow(HMDA_female_Recovery), floor(.50*nrow(HMDA_female_Recovery)), replace = F)
train<-HMDA_female_Recovery[sample.data, ]
test<-HMDA_female_Recovery[-sample.data, ]

##predicted probabilities for test data based on training data
preds<-predict(model,newdata=test, type="response")

##see the predicted probabilities for the test data and their classification based on threshold of 0.5
display<-test[,c(1,4,5,6,8,13,14,15,16,17)]
display<-cbind(display,preds,preds>0.5)
names(display)[12]<-"ClassYes"
display

##confusion matrix when threshold is 0.5. True values in the rows. 
table(test$action_taken_name,preds > 0.5)

##need ROCR package to produce ROC curve
library(ROCR)

##produce the numbers associated with classification table
rates<-ROCR::prediction(preds, test$action_taken_name)

##store the true positive and false postive rates
roc_result<-ROCR::performance(rates,measure="tpr", x.measure="fpr")

##compute the AUC
auc<-ROCR::performance(rates, measure = "auc")
auc@y.values

# Accuracy
(301 + 1219) / (301 + 687 + 169 + 1219)

# Test Error Rate
1 - 0.6397306 #0.3603 = 36.03%

# FPR
687 / (301 + 687) #0.6953 = 69.53%

# FNR
169 / (169 + 1219) #0.1218 = 12.18%

```

# Answer to question: Are white female applicants more likely to be approved than black female applicants and asian female applicants in the recovery period?

We get stat. significant intercepts for app_raceAsian(0.9224) and app_raceWhite(0.9276) with the reference class being Black. This means that the estimated odds of loan approval for asian females is exp(0.9224) = 2.5153 times the estimated odds of approval for black females, given other factors affecting approval are held constant. Moreover, the estimated odds of loan approval for white females is exp(0.9276) = 2.5284 times the estimated odds of approval for black females given other factors are held constant. This result is in alignment with the fact that black applicants are lesser likely to be approved than asians and white applicants which was expected given the literature review.

# Answer to question: Are female residents of CA more likely to be approved for a mortgage than VA residents in the recovery period?

To find the The estimated odds of loan approval for female VA residents as compared to the estimated odds of approval for female CA residents, given other factors are held constant, we also have to take into account the inetraction between state and loan to income ratio. Since CA is the reference class, this means that the difference between VA and CA would be the sum of coeffients for state_abbr and the coefficient for the interaction between state_abbr and loan_to_income. This means that the estimated odds of loan approval for VA applicants is exp(-5.837 - 0.2673) = 0.002233244 times that for CA, given other factors are held constant. However, this is an insignificant result, showing that state doesn't play a huge role in determining approval for female applicants in the recovery period.

## Now doing the same for the recession period also

## Subsetting for females and recession
```{r}
# Step 1: Subsetting the data
HMDA_female_Recession <- HMDA %>%
  filter(period == "Recession" & applicant_sex_name == "Female")
```


```{r}
#Step 2 - additional EDA for subsetted data of females and recession

#EDA - for female applicants and the Recession period (for interaction terms)

## exploring the relation for interactions

#i) interaction between loan_to_income and property_type_name
ggplot(HMDA_female_Recession, aes(x = property_type_name, y = loan_to_income, fill = action_taken_name)) +
  geom_violin(scale = "width", trim = FALSE) +
  labs(title = "Female Loan to Income Ratio by Property Type and Loan Status (Recession)",
       x = "Property Type",
       y = "Loan to Income Ratio",
       fill = "Loan Status") +
  theme_minimal() +
  scale_fill_manual(values = c("Loan originated" = "blue", "Loan denied" = "red"))


##ii) interaction between loan_to_income and state_abbr
ggplot(HMDA_female_Recession, aes(x = state_abbr, y = loan_to_income, fill = action_taken_name)) +
  geom_violin(scale = "width", trim = FALSE) +
  labs(title = "Female Loan to Income Ratio by State and Loan Status (Recession)",
       x = "State",
       y = "Loan to Income Ratio",
       fill = "Loan Status") +
  theme_minimal() +
  scale_fill_manual(values = c("Loan originated" = "blue", "Loan denied" = "red"))

##iii) interaction between loan_to_income and loan_purpose_name
ggplot(HMDA_female_Recession, aes(x = loan_purpose_name, y = loan_to_income, fill = action_taken_name)) +
  geom_violin(scale = "width", trim = FALSE) +
  labs(title = "Female Loan to Income Ratio by Loan Purpose and Loan Status (Recession)",
       x = "Loan Purpose",
       y = "Loan to Income Ratio",
       fill = "Loan Status") +
  theme_minimal() +
  scale_fill_manual(values = c("Loan originated" = "blue", "Loan denied" = "red"))

##iv) interaction between per capita GDP and CPI
ggplot(HMDA_female_Recession, aes(x = Per_Cap_GDP, y = CPI, color = action_taken_name)) +
  geom_smooth(method = "lm", se = F) +
  labs(title = "State-Level Per Capita GDP vs. CPI (Recession)",
       x = "Per Capita GDP",
       y = "CPI",
       color = "Loan Status") +
  theme_minimal()

```

v) State and Approval Rate

```{r}
# Calculate approval rate
approval_data_recession <- HMDA_female_Recession %>%
  group_by(state_abbr, loan_type_name, app_race, action_taken_name) %>%
  summarize(count = n()) %>%
  tidyr::spread(action_taken_name, count, fill = 0) %>%
  mutate(approval_rate = `Loan originated` / (`Loan originated` + `Loan denied`) * 100)

approval_data_summary <- approval_data_recession %>%
  group_by(state_abbr) %>%
  summarize(mean_approval_rate = mean(approval_rate, na.rm = TRUE))

ggplot(approval_data_summary, aes(x=state_abbr, y=mean_approval_rate)) +
  geom_bar(stat="identity", fill="blue") +
  labs(title="Female Mean Approval Rate by State (Recession)",
       x="State", 
       y="Mean Approval Rate (%)") + 
  theme_minimal() +
  scale_y_continuous(limits = c(0,60,20))
```
vi) Loan Type and Approval Rate
```{r}
approval_data_summary <- approval_data_recession %>%
  group_by(loan_type_name) %>%
  summarize(mean_approval_rate = mean(approval_rate, na.rm = TRUE))

ggplot(approval_data_summary, aes(x=loan_type_name, y=mean_approval_rate)) +
  geom_bar(stat="identity", fill="blue") +
  labs(title="Female Mean Approval Rate by Loan Type (Recession)",
       x="Loan Type", 
       y="Mean Approval Rate (%)") + 
  theme_minimal() +
  scale_y_continuous(limits = c(0,100,20))
```
vii) Race and Approval Rate
```{r}
approval_data_summary <- approval_data_recession %>%
  group_by(app_race) %>%
  summarize(mean_approval_rate = mean(approval_rate, na.rm = TRUE))

ggplot(approval_data_summary, aes(x=app_race, y=mean_approval_rate)) +
  geom_bar(stat="identity", fill="blue") +
  labs(title="Female Mean Approval Rate by Race (Recession)",
       x="Race", 
       y="Mean Approval Rate (%)") + 
  theme_minimal() +
  scale_y_continuous(limits = c(0,80,20))
```
ii) Logistic Regression for Recession 

```{r}
# Step 3: Logistic Regression

# Having Black as the reference class
HMDA_female_Recession$app_race <- relevel(HMDA_female_Recession$app_race, ref = "Black")

model <- glm(action_taken_name ~ state_abbr + loan_type_name + property_type_name +
             loan_purpose_name + loan_to_income + app_race + CPI + unemployment_rate + Per_Cap_GDP + loan_to_income*property_type_name + loan_to_income*state_abbr + loan_to_income * loan_purpose_name + Per_Cap_GDP * CPI, data = HMDA_female_Recession, family = binomial)

# Step 4: Hypothesis Testing
# Get a summary of the model to check p-values
summary(model)

##set seed so results are reproducible
set.seed(111)

##evenly split data into train and test sets
sample.data<-sample.int(nrow(HMDA_female_Recession), floor(.50*nrow(HMDA_female_Recession)), replace = F)
train<-HMDA_female_Recession[sample.data, ]
test<-HMDA_female_Recession[-sample.data, ]

##predicted probabilities for test data based on training data
preds<-predict(model,newdata=test, type="response")

##see the predicted probabilities for the test data and their classification based on threshold of 0.5
display<-test[,c(1,4,5,6,8,13,14,15,16,17)]
display<-cbind(display,preds,preds>0.5)
names(display)[12]<-"ClassYes"
display

##confusion matrix when threshold is 0.5. True values in the rows. 
table(test$action_taken_name,preds > 0.5)

##need ROCR package to produce ROC curve
library(ROCR)

##produce the numbers associated with classification table
rates<-ROCR::prediction(preds, test$action_taken_name)

##store the true positive and false postive rates
roc_result<-ROCR::performance(rates,measure="tpr", x.measure="fpr")

##compute the AUC
auc<-ROCR::performance(rates, measure = "auc")
auc@y.values

# Accuracy
(1475 + 708) / (1475 + 508 + 792 + 708)

# Test Error Rate
1 - 0.6267585 #0.3732 = 37.32%

# FPR
508 / (1475 + 508) #0.2562 = 25.62%

# FNR
792 / (792 + 708) #0.528 = 52.8%
```
```{r}
exp(0.7709)
```

# Answer to question: Are white female applicants more likely to be approved than black female applicants and asian female applicants in the recession period?

We get stat. significant intercepts for app_raseAsian(0.9859) and app_raceWhite(0.7709) with the reference class being Black. This means that the estimated odds of loan approval for asian females is exp(0.9859) = 2.6802 times the estimated odds of approval for black females, given other factors affecting approval are held constant. Moreover, the estimated odds of loan approval for white females is exp(0.7709) = 2.1617 times the estimated odds of approval for black females given other factors are held constant. This result is in alignment with the fact that during recession, black female applicants are still less likely to be approved than white females and asian females. However, the racial gap between black females and white females decreases which is alignment with the fact that black applicants had greater accessibility to home loans during recovery; however, they had access to subprime mortgages which make then financially more vulnerable. However, I don't know why the racial gap has increased between asian and black females. In fact, asian females are the most likely to be approved for a mortgage.

# Answer to question: Are female residents of CA more likely to be approved for a mortgage than VA residents in the recession period?

We get a statistically insignificant result saying that the estimated odds of loan approval for female VA residents is exp(1.228 - 0.0420) = 3.2740 times the estimated odds of approval for female CA residents, given other factors are held constant. This is an insignificant significant result which isn't in alignment with the literature review saying that the mortgage market in CA was one of the worst affected by the great recession, or the fact that CA is a more loiberal state than VA.

# Conclusion

# Q1) Are white female applicants more likely to be approved than black female applicants and asian female applicants in the recession period? What about the recovery period?

During the recession period,

We get stat. significant intercepts for app_raseAsian(0.9859) and app_raceWhite(0.7709) with the reference class being Black. This means that the estimated odds of loan approval for asian females is exp(0.9859) = 2.6802  times the estimated odds of approval for black females, given other factors affecting approval are held constant. Moreover, the estimated odds of loan approval for white females is exp(0.7709) = 2.1617 times the estimated odds of approval for black females given other factors are held constant. This result is in alignment with the fact that during recession, black female applicants are still less likely to be approved than white females and asian females. However, the racial gap between black females and white females decreases which is alignment with the fact that black applicants had greater accessibility to home loans; however, they had access to subprime mortgages which make then financially more vulnerable. However, I don't know why the racial gap has increased between asian and black females. In fact, asian females are the most likely to be approved for a mortgage.

During the recovery period,

We get stat. significant intercepts for app_raceAsian(0.9224) and app_raceWhite(0.9276) with the refernce class being Black. This means that the estimated odds of loan approval for asian females is exp(0.9224) = 2.5153 times the estimated odds of approval for black females, given other factors affecting approval are held constant. Moreover, the estimated odds of loan approval for white females is exp(0.9276) = 2.5284 times the estimated odds of approval for black females given other factors are held constant. This result is in alignment with the fact that black applicants are lesser likely to be approved than asians and white applicants which was expected given the literature review.

# Q2) Are female residents of CA more likely to be approved for a mortgage than female VA residents in the receccion period? What about the recovery period?

This is insignificant result (for Recession and Recovery), showing that state doesn't play a huge role in determining approval for female applicants in the recovery period.
